Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-26 paper/paper.Rmd
Intro updated

---
title: "Distortion-corrected kernel density estimate on Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = T,
  messages = FALSE,
  warning = FALSE,
  eval = TRUE
  # include = FALSE
)
options(tinytex.verbose = TRUE)
library(tidyverse)
# library(dimRed)
# library(reticulate)
library(here)
# library(viridis)
library(hdrcde)
# library(igraph)
# library(matrixcalc)
# library(akima)
library(car)
# library(ggforce)
# library(ks)
library(patchwork)
# library(copula)
# library(plotly)
library(kableExtra)
# Jmisc::sourceAll(here::here("R/sources"))
# set.seed(1234)
```

@@

<!-- Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data.  -->
Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data.
It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data. In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies. To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates.

Kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] is one of the most popular methods to calculate the probability density function from a sample dataset. KDE is flexible to learn the shape of the underlying density from the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut]. There is extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp]. For multivariate data, variable kernel density estimator [VKDE; @Jones1990-oe; @Terrell1992-ut] has been studied with an adaptive bandwidth matrix to control the amount of smoothing on the location of the estimated point [balloon estimator; @Terrell1992-ut] or the data point [sample smoothing estimator; @Terrell1992-ut]. However, these density estimators are based on random samples in the Euclidean space.

For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu proposed a kernel density estimator based on the kernel weight depending on the distance between the estimated points and the sample data points. The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature. [@Henry2009-ll] In many application scenarios, we tend to find that the sample points on a manifold are embedded in a much higher-dimensional space, and the kernel density estimator from @Pelletier2005-vu is not directly applicable.
This is when we introduce manifold learning to reduce the input data dimension. For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), are applied to get a low-dimensional embedding. We propose a kernel density estimator to be applied to the low-dimensional embedding and define outliers as those with the lowest densities.

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold [@Goldberg2008-co]. To tackle this problem and measure the distortion incurred in manifold learning,
@Perrault-Joncas2013-pq propose the Learn metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.

The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding. Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others. The Riemannian gives us an idea of the direction and angle of this stretching, which is informative for learning the manifold. In variable kernel density estimate, the bandwidth matrix $H$ is also defined to control the amount of smoothing for each data point.
Therefore, we propose to the use Riemannian metric as the bandwidth matrix in, we could further get the kernel density estimation of the manifold $M$. This kernel density estimate can then be used to produce the highest density region plots[@Hyndman1996-lk] for outlier visualization.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds. Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator to detect anomalies based on the embedding. \autoref{simulation} is composed of two simulations with the proposed algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D swiss roll structure and the second with a 5-D semi-hypersphere mapped in a 100-D space. Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use the Appendix in \autoref{riemgeo} as a reference.

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data. @Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data.
In general, given multivariate random samples $\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d$
<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathbb{R}^d$, $i = 1, 2, \dots, N$  -->
drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x^1, x^2, \dots, x^d)^\prime \in \mathbb{R}^d$ is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
\end{equation}
where $K(\pmb{x})$ is the kernel function, $\pmb{H}_i$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})$.
When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function. Then an average over all data points is taken as the estimated density $\hat{f}$. KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point $\pmb{x}$.

@@

(ref:vkdecaption) The proposed schematic for variable kernel density estimation with recovered geometry.

>>>>

---
title: "Distortion-corrected kernel density estimate on Riemannian manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
abstract: "Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel distortion-corrected kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data."
keywords: "manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: false
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = T,
  messages = FALSE,
  warning = FALSE,
  eval = TRUE
  # include = FALSE
)
options(tinytex.verbose = TRUE)
library(tidyverse)
library(here)
# library(hdrcde)
# library(car)
library(patchwork)
library(kableExtra)
# Jmisc::sourceAll(here::here("R/sources"))
# set.seed(1234)
```

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data which usually belongs to a subset of $\mathbb{R}^d$. Applications of kernel density estimation [KDE; @Parzen1962-gt; @Chen2017-dw] include finding hot spots of traffic network in the GIS environment [@Xie2008-eb; @Okabe2009-nb], automatic detection in visual surveillance systems [@Elgammal2002-cw], wind power density detection [@Jeon2012-ac], prime prediction via Twitter messages [@Gerber2014-tq], and so on. However, when samples are assumed to be drawn from a Riemannian manifold embedded in a high-dimensional space of much more than $\mathbb{R}^d$, kernel density estimation has to be generalized to a non-Euclidean space and further approximation methods have to be adapted. @Pelletier2005-vu propose a kernel density estimator based on the Riemannian geodesic distance of the manifold but it is only applicable when the underlying manifold is known. Manifold learning algorithms could be applied to reduce the dimension and get an approximation of the manifold, but different manifold learning algorithms could induce different distortions of the same manifold. Therefore, we propose a distortion-corrected kernel density estimator for Riemannian manifolds embedded in more than $\mathbb{R}^d$. Our estimator could be applied to any reasonable manifold learning embedding from the high-dimensional sample data and fix the distortions at each point with estimated Riemannian geodesic distance and volume density function. This estimator could be further applied for unsupervised tasks such as anomaly detection where the outliers are the lowest density points.

<!-- Anomaly detection has been an important and diverse area where anomalies or outliers are detected in a given data. It often involves a data analysis process to uncover the unusual patterns and has been widely applied to machine learning [@Omar2013-lz], network intrusions identification [@Ahmed2016-co; @Bhuyan2013-ra], medical imaging [@Fernando2022-ef], fraudulent attacks [@Ahmed2016-dz], cyber-security [@Ten2011-co], energy consumption [@Cheng2021-ex], the last of which explored detecting households with anomalous electricity usage distributions instead of raw data. In the case of non-Euclidean sample space, the observations lie on a Riemannian manifold embedded in a very high-dimensional ambient space, which makes it computationally expensive or impossible to detect anomalies. To address this problem, we propose a kernel density estimator of the low-dimensional statistical manifold embedding and find outliers as the data with the lowest density estimates. -->

For a given kernel function, kernel density estimation is flexible to learn the shape of the underlying density of the data controlled by the bandwidth and the selection of bandwidth is crucial in KDE [@Jones1990-oe; @Terrell1992-ut]. 
<!-- There are extensive research on bandwidth selection, two main categories of which are cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [@Wand1994-gc; @Duong2003-sp].  -->
Many bandwidth selection methods have been proposed in the literature, including the rule-of-thumb, cross-validation [@Jones1992-ta; @Sain1994-gr] and plug-in methods [See @Heidenreich2013-bl; @Scott2015-vl for details]. 
For univariate kernel density estimation, the bandwidth selection problem has been thoroughly investigated [See @Jones1992-ef; @Cao1994-st; @Jones1996-cb; @Wand1994-xu for reviews]. The generalization to multivariate case could mostly be found in @Duong2003-sp, @Duong2004-rh, and @Chacon2010-wm. In this paper, we focus on the multivariate kernel density estimation.

Note that a fixed bandwidth matrix $\pmb{H}$ is a global smoothing parameter for all data points. However, when the local data structure is not universal for all sample data, which is true in most applications, an adaptive bandwidth matrix that is varying rather than fixed at each data point is needed. 
The bandwidth is varied depending on either the location of the sample points [sample smoothing estimator; @Terrell1992-ut] or that of the estimated points [balloon estimator; @Terrell1992-ut]. In this paper, the densities are estimated at the sample points themselves, so we only need to consider the case where the bandwidth changes for each sample point and will refer to this as the *variable/adaptive kernel density estimation* [VKDE; Section 6.6 of @Scott2015-vl] unless otherwise stated.
However, these kernel density estimators are based on random samples in the Euclidean space.

For samples points lying on a manifold with the differentiable structure called the Riemannian manifold, @Pelletier2005-vu generalize the kernel density estimator based on the kernel weights from the geodesic distance between the estimated points and the sample points. The idea of the estimator is to use a strictly positive function of the geodesic distance on the manifold and then normalize it with the volume density function of the Riemannian manifold for curvature [@Henry2009-ll]. 
However, in many application scenarios, we tend to find that the sample points are not drawn directly from the manifolds because they are embedded in a much higher-dimensional space. Therefore, the kernel density estimator from @Pelletier2005-vu is not applicable because the geodesic distance and the volume density function are unknown.
This is when we introduce manifold learning to reduce the input data dimension. For these high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP (see details of these algorithms in @Cheng2021-ex), could be applied to get a low-dimensional embedding, which are used as approximations of the underlying manifold.

In manifold learning, the underlying idea is that the data lies on a low-dimensional smooth manifold that is embedded in a high-dimensional space. One of the fundamental objectives of manifold learning is to explore the geometry of the dataset, including the distances between points and volumes of regions of data. These intrinsic geometric attributes of the data, such as distances, angles, and areas, however, can be distorted in the low-dimensional embedding, leading to failure to recover the geometry of the manifold [@Goldberg2008-co]. To tackle this problem and measure the distortion incurred in manifold learning,
@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any existing embedding output with geometric information in the Riemannian metric of the manifold itself. By applying this algorithm, the outputs of different manifold learning methods can be unified and compared under the same framework, which would highly benefit in improving the effectiveness of the embedding.
The Riemannian metric using the method of @Perrault-Joncas2013-pq gives some idea of the distortion of an embedding. Mapping the points through a non-linear function "stretches" some regions of space and "shrinks" others. The Riemannian gives us an idea of the direction and angle of this stretching at each point, which is informative for learning the manifold.

By exploiting the connection between the estimated Riemannian metric and the Riemannian geodesic distance as well as the volume density function for curvature, we propose the main contribution of the paper, which is the variable distortion-corrected kernel density estimator (DC-KDE) for manifold learning embedding. Starting from the high-dimensional sample data, we apply manifold learning algorithms to get the low-dimensional embedding in the same dimensional space as the underlying manifold together with the estimated Riemannian matrix at each embedding point. Then the DC_KDE is used to estimate the density of the manifold and distortion induced by manifold learning methods are fixed with the estimated geometric information. Our distortion-corrected KDE is novel in filling the gap between the high-dimensional sample space and the density of the unknown manifold. These density estimates are useful in many other areas, including classification, clustering and anomaly detection. Similar to @Cheng2021-ex, the highest density region plots[@Hyndman1996-lk] could be generated using the kernel density estimates for outlier visualization, which brings a novel anomaly detection method for Riemannian manifolds.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. We start by introducing the multivariate kernel density estimate with adaptive bandwidth and the kernel density estimator for Riemannian manifolds. Then we provide justification for the use of Riemannian metric to correct the distortions in manifold learning embedding and further apply the proposed estimator for anomaly detection. \autoref{simulation} is composed of two simulations with the proposed anomaly detection algorithm; the first deals with 2-dimensional data from gaussian mixture model mapped into a 3-D swiss roll structure and the second with a 5-D semi-hypersphere data mapped in a 100-D space. Different manifold learning algorithms are applied to the high-dimensional data to get the low-dimensional embedding which are then used to estimate densities and detect anomalies.
\autoref{application} contains the application to visualize and identify anomalous households in the Irish smart meter dataset. Conclusions and discussions are presented in \autoref{conclusion}. Readers interested in the notions of Riemannian geometry mentioned in this paper could use \autoref{riemgeo} as a reference.

@@

Kernel density estimation is one of the most popular methods to calculate the probability density function from a sample data. @Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data.
In general, given multivariate random samples $\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d$
<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathbb{R}^d$, $i = 1, 2, \dots, N$  -->
drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x^1, x^2, \dots, x^d)^\prime \in \mathbb{R}^d$ is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
\end{equation}
where $K(\pmb{x})$ is the kernel function, $\pmb{H}_i$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})$.
When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function. Then an average over all data points is taken as the estimated density $\hat{f}$. KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point $\pmb{x}$.

@@

(ref:vkdecaption) The proposed schematic for anomaly detection with distortion-corrected kernel density estimates.
